---
title: "Preparation of IBDMDB sequencing metadata"
subtitle: "Integration of ENA and NCBI SRA run-level metadata for metagenomic network and pathway analyses"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
editor: source
---

# Aim of the notebook

This notebook prepares the sequencing metadata used for the analysis of the
Inflammatory Bowel Disease Multi'omics Database (IBDMDB) dataset.

The main goal is to build a clean and consistent run-level metadata table by
integrating information retrieved from two complementary public resources:

1. the ENA read-run file report for BioProject `PRJNA398089`;
2. the NCBI SRA Run Selector table for the same BioProject.

The resulting table will be used to select the subset of samples relevant for
the downstream metagenomic analyses, with particular focus on shotgun
metagenomic data suitable for microbial community, network, and pathway-based
analyses.

# Data sources

The starting metadata files are:

- `filereport_read_run_PRJNA398089.tsv`, downloaded from the ENA Portal API;
- `SraRunTable_PRJNA398089.csv`, downloaded from the NCBI SRA Run Selector.

Both files describe sequencing runs associated with the IBDMDB/iHMP study, but
they contain partially overlapping and partially complementary information.
For this reason, they are first standardized and compared, and then joined into
a single metadata table.

# General workflow

The notebook follows these steps:

1. load the ENA and SRA metadata tables;
2. standardize column names from the SRA table;
3. harmonize key identifiers and technical fields across the two sources;
4. check consistency of run accessions and shared columns;
5. merge ENA and SRA metadata into a single table;
6. clean selected clinical and technical variables;
7. filter the metadata to retain the samples suitable for the downstream
   metagenomic analysis.

```{r setup, message=FALSE, warning=FALSE}
# Load packages used for data manipulation and metadata cleaning.
library(tidyverse)
library(janitor)
```

```{r, message=FALSE, warning=FALSE}
report_ena <- read_tsv("../raw/filereport_read_run_PRJNA398089.tsv") %>%
  rename(center_code = center_name)
metadata_sra <- read_csv("../raw/SraRunTable_PRJNA398089.csv")
```

```{r}
colnames(report_ena)
```

```{r, message=FALSE}
metadata_sra_std <- metadata_sra %>%
  janitor::clean_names()
colnames(metadata_sra_std)
```

```{r}
metadata_sra_std <- metadata_sra_std %>%
  rename(
    run_accession = run,
    library_strategy = assay_type,
    base_count = bases,
    study_accession = bio_project,
    sample_accession = bio_sample,
    experiment_accession = experiment,
    instrument_model = instrument,
    library_layout = library_layout,
    library_source = library_source,
    center_name = center_name,
    sample_alias = sample_name,
    study_accession_sra = sra_study,
    scientific_name = organism,
    instrument_platform = platform
  )
```

```{r}
all(report_ena$run_accession %in% metadata_sra_std$run_accession)
all(metadata_sra_std$run_accession %in% report_ena$run_accession)

report_ena |>
  count(run_accession) |>
  filter(n > 1)

metadata_sra_std |>
  count(run_accession) |>
  filter(n > 1)
```

## Check consistency of shared columns

After harmonizing the main column names, the two metadata tables are compared
using the columns they have in common.

```{r}
shared_columns <- intersect(colnames(report_ena), colnames(metadata_sra_std))
```

```{r}
report_ena_check <- report_ena %>%
  arrange(run_accession) %>%
  select(all_of(shared_columns))

metadata_sra_std_check <-  metadata_sra_std %>%
            arrange(run_accession) %>%
            select(all_of(shared_columns))
```

```{r}
identical(report_ena_check, metadata_sra_std_check)
```

## Join ENA and SRA metadata

The ENA and SRA tables are joined using the run accession and the main shared
technical identifiers.

```{r}
metadata_join <- report_ena %>%
  left_join(metadata_sra_std, 
            by = join_by(run_accession, study_accession, sample_accession, 
                         experiment_accession, scientific_name,
                         instrument_platform, instrument_model, library_layout,
                         library_strategy, library_source, base_count, sample_alias))
```

# Reorder metadata columns

The joined metadata table is reorganized to make it easier to inspect and use.
Columns are grouped into conceptual blocks:

- main sequencing and sample identifiers;
- subject-level and clinical information;
- biological and environmental sample descriptors;
- library and sequencing information;
- sequencing magnitude and file size;
- FASTQ download information;
- release, storage, and provenance metadata.

```{r}
metadata_join <- metadata_join %>%
  relocate(
    # main IDs
    run_accession,
    sample_alias,
    external_sample_id,
    sample_accession,
    experiment_accession,
    study_accession,
    study_accession_sra,

    # subject/clinical info
    host_subject_id,
    host_disease,
    host_sex,
    host_age,
    visit_num,
    replicate,
    collection_date,
    consent,

    # bioinformatic/technical sample info
    scientific_name,
    tax_id,
    host,
    isolation_source,
    geo_loc_name,
    geo_loc_name_country,
    geo_loc_name_country_continent,
    lat_lon,
    env_biome,
    env_feature,
    env_material,

    # library and sequencing
    library_strategy,
    library_layout,
    library_source,
    library_selection,
    library_name,
    instrument_platform,
    instrument_model,
    center_name,
    center_code,
    experiment_title,

    # magnitude
    read_count,
    base_count,
    avg_spot_len,
    bytes,
    fastq_bytes,

    # Download
    fastq_ftp,
    fastq_md5,

    # date / storage
    release_date,
    create_date,
    datastore_filetype,
    datastore_provider,
    datastore_region,
    bio_sample_model,
    version
  )
```

# Standardize disease labels

Disease labels are harmonized to obtain a consistent set of values across the
metadata table.

The original metadata contain different representations of the same condition,
such as abbreviations and full names. These values are converted to compact
snake-case labels that are easier to use in downstream analyses.

```{r}
metadata_join <- metadata_join %>% 
  mutate(
    host_disease = if_else(is.na(host_disease), "missing", host_disease),
    host_disease = str_to_lower(host_disease),
    host_disease = case_when(
      host_disease %in% c("cd", "crohn''s disease", "crohn's disease", "crohn’s disease") ~ "crohns_disease",
      host_disease %in% c("uc", "ulcerative colitis") ~ "ulcerative_colitis",
      host_disease %in% c("nonibd", "non-ibd") ~ "control",
      TRUE ~ host_disease
    ))
```

```{r}
metadata_join %>%
  count(host_disease, sort = TRUE)
```
# Select shotgun metagenomic sequencing runs

The full metadata table contains multiple assay types, including shotgun
metagenomic, metatranscriptomic, viral metagenomic, and amplicon sequencing
data.

For the main downstream analysis, only shotgun metagenomic runs from human gut
metagenome samples are retained.

Before filtering, the available combinations of library strategy, scientific
name, library selection, and library source are summarized.

```{r}
metadata_join %>%
  count(library_strategy, scientific_name, library_selection, library_source, sort = TRUE)
```

```{r}
metadata_join_filt <- metadata_join %>%
  filter(library_strategy == "WGS",
         scientific_name == "human gut metagenome",
         library_selection == "RANDOM",
         library_source == "METAGENOMIC")
```

```{r}
metadata_join_filt %>%
  arrange(read_count) %>%
  select(run_accession, host_subject_id, host_disease, read_count, base_count, avg_spot_len) %>%
  print(n = 20)
```

```{r}
metadata_join_filt <- metadata_join_filt %>%
  filter(read_count >= 1e5)
```


## Inspect missing disease labels at subject level

Disease status should be consistent within each subject. Therefore, subjects
with multiple disease labels are inspected.

In particular, this step checks whether any subject has both a missing disease
label and a non-control disease label. Such cases require manual inspection
before assigning missing values to the control group.

```{r}
metadata_join_filt %>%
  group_by(host_subject_id, host_disease) %>%
  count() %>%
  ungroup() %>%
  group_by(host_subject_id) %>%
  filter(n() > 1) %>%
  arrange(host_subject_id) %>%
  group_by(host_subject_id) %>%
  filter(
    any(host_disease == "missing"),
    !any(host_disease == "control"))
```

## Build the final metadata table

The final metadata table is generated by applying a set of inclusion criteria
for the downstream analysis.

The filtering steps are:

1. remove the subject/run combination with ambiguous missing disease annotation;
2. retain adult subjects only;
3. convert the remaining missing disease annotations to `control`;
4. retain subjects with at least 10 metagenomic sequencing runs.

The last criterion is useful for longitudinal or repeated-measure analyses,
where subjects with too few observations may not provide enough information to 
reconstruct the correlation network.

```{r}
metadata_join_final <- metadata_join_filt %>%
  filter(!(host_subject_id == 4004 & host_disease == "missing")) %>%
  filter(host_age >= 18) %>%
  mutate(host_disease = if_else(host_disease == "missing", "control", host_disease)) %>%
  group_by(host_subject_id) %>%
  filter(n() >= 10) %>%
  ungroup() 
```

```{r}
metadata_join_final %>%
  summarise(
    n_runs = n(),
    n_subjects = n_distinct(host_subject_id),
    .by = host_disease
  ) %>%
  arrange(desc(n_runs))
```


```{r}
metadata_join_final %>%
  group_by(host_subject_id, host_disease) %>%
  count() %>%
  arrange(n)
```

```{r}
metadata_join_final %>%
  distinct(host_subject_id, host_disease) %>%
  group_by(host_subject_id, host_disease) %>%
  count() %>%
  arrange(n)
```


```{r}
write_tsv(metadata_join_final, "../processed/ibdmdb_metadata_runs_mgx.tsv")
```

